---
title: "Activity 1"
author: "David Dominguez - 29092830"
date: "2025-02-14"
output: html_document
---

## Librerias

```{r Librerias}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(readxl)
library(corrplot)
library(rpart)
library(rpart.plot)
library(stringr)
library(stringi)
library(janitor)
library(lubridate)
library(purrr) 
```

## Dataset

### Glosario:
**new_fdi:**	New Foreign Direct Investment Inflows. Millions of Dollars. 
**reinv_profits:**	Reinvestment of Profits - Foreign Direct Investment Inflows. Millions of Dollars. 
**intercom_acc:**	Intercompany Accounts - Foreign Direct Investment Inflows. Millions of Dollars. 
**total_fdi:**	Total Foreign Direct Investment Inflows. Millions of Dollars. 
**crime_rate:**	Crime rate per 100,000 state's population. 
**unemployment:**	Percentaje of unemployed population. 
**employment:** 	Percentaje of employed population. 
**business_activity:**	Index of economic activity weighted by the state's distance to nearest U.S. port of entry. 
**real_wage:**	Real wage using 2018 INPC = 100. MXN Pesos.
**pop_density:**	Population per state's land area km2. 
**good_governance:**	Ratio between state's public investment and its public debt. 
**ratio_public_investment:**	Ratio between state's public investment and its gross domestic product (GDP). 
**lq_primary:**	Location quotient of employed people in primary industry. 
**lq_secondary:**	Location quotient of employed people in secondary sector. 
**lq_tertiary	Location:** quotient of employed people in tertiary sector. 
**exchange_rate:**	MXN pesos per 1 USD
**patents_rate:**	Number of R&D patentes per 100,000 state's population
**inpc:**	National Price Consumer Index. 2018 = 100

```{r Dataset, echo=FALSE, message=FALSE, warning=FALSE}
nearshoring <- read_excel('panel_data.xlsx')

# Renombrar columnas
nearshoring <- nearshoring %>%
  rename(region_name = `region...23`, 
         region_num = `region...24`)

# Verificar los cambios
colnames(nearshoring)
```

## Overview and Adjustments

b. Crear y/o transformar variables (e.g., variables nominales, variables reales, variables ponderadas, etc.)

```{r Modifications, message=FALSE, warning=FALSE}
# Convertir valores de USD a MXN
nearshoring <- nearshoring %>%
  mutate(across(c(new_fdi, reinv_profits, intercom_acc, total_fdi, exports_miles_dlls), 
                ~ . * exchange_rate, .names = "mxn_{.col}"))

# Ajustar valores a precios reales (2018)
nearshoring <- nearshoring %>%
  mutate(across(c(mxn_new_fdi, mxn_reinv_profits, mxn_intercom_acc, mxn_total_fdi, real_wage, mxn_exports_miles_dlls), 
                ~ . / inpc * 100, .names = "real_{.col}"))

# Subset solo con variables reales en MXN (sin dólares)
nearshoring_clean <- nearshoring %>%
  select(state, year, region_name, region_num, 
         real_mxn_new_fdi, real_mxn_reinv_profits, real_mxn_intercom_acc, 
         real_mxn_total_fdi, real_wage, real_mxn_exports_miles_dlls, 
         crime_rate, unemployment, employment, business_activity, 
         pop_density, good_governance, ratio_public_investment, 
         lq_primary, lq_secondary, lq_tertiary, patents_rate, 
         exchange_rate, inpc, border_distance, trump_election)

# Verificar cambios
summary(nearshoring_clean)

# Visualizar primeras filas
head(nearshoring_clean)
```


```{r}

```


## Analisis Exploratorio

### Estadísticos Descriptivos y de Dispersión

### Visualización de los Datos

#### Análisis Univariado



#### Análisis Bivariado



## Hallazgos



## Referencias

